---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This Data analysis is created for a peer assessment for  
[Reproductive Reserarch](https://www.coursera.org/course/repdata) course which is module in 
Data Science Specialization by Johns Hopkins University.

Author: Marcin Wiśniowski

Date: 2015-01-14

Libraries used for assessment:
```{r libraries, message=FALSE}
library("data.table")
library("ggplot2")
library("lubridate")
```

## Dataset

The data for this assignment can be downloaded from the course website:
[Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 52K

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD format

* **interval**: Identifier for the 5-minute interval in which measurement was taken


The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 
observations in this dataset.

## Loading and preprocessing the data
Dataset should be stored in the same directory as processing script file. First step is to check if
Dataset file is available and then uncompress it for future processing.
```{r getting_data}
if( !file.exists("activity.zip") ){
    download.file(
        "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
        destfile="activity.zip",
        method="curl"
        )
}
unzip("activity.zip")
```

Unzipped file can be easily read with read.csv function. First step is to transform date coded as
YYYY-MM-DD to the date object within data table.

```{r reading_data}
data <- fread("activity.csv")
data$date <- ymd(data$date)
```

## What is mean total number of steps taken per day?

During this analysis the missing values will be ignored. As a preparation for this task new data
table will be created. This data table will have two columns:

* **date**: The date on which the measurement was taken coded as POSIXct date object

* **steps**: Total number of step that were mesured through day

```{r total_steps_per_day}
data.total_steps <- data[,sum(steps, na.rm=TRUE),by=date]
setnames(data.total_steps, c("date", "steps"))
```

This table can be easily visualized as a histogram

```{r plot_total_steps_per_day}
plot.total_steps <- ggplot( data.total_steps, aes(x=steps) ) 
plot.total_steps <- plot.total_steps + geom_histogram(binwidth = 1000, aes(fill = ..count..))
plot.total_steps <- plot.total_steps + xlab("Total steps per day") + ylab("Frequency")
plot.total_steps <- plot.total_steps + ggtitle("Frequency distribution of total steps per day")
plot.total_steps
```

Analyzing the raw nmbers, it gives the conclusion of:

* Mean number of steps per day: **9354**

* Median of steps per day: **10395**

```{r summary_total_steps_per_day}
summary( data.total_steps )
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
